Evolutionary Learning of Policies for MCTS Simulations

### 2012-03-24 thesis/paper/paper.tex
Using acm_proc_article-sp style, had to tweak the figures and itemize/enumerate environments to fit into 8 pages.

\documentclass{acm_proc_article-sp}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{url}
\usepackage{havannah}
\newcommand{\hblack}{\texttt{Black}}
\newcommand{\hwhite}{\texttt{White}}
\newcommand{\hblacks}{\texttt{Black's}}
\newcommand{\hwhites}{\texttt{White's}}
\newcommand{\loc}[1]{\texttt{#1}}

@@

@@

@@

@@

\section{Introduction}
Abstract strategy games provide an attractive testbed for artificial intelligence and machine learning applications. The game framework provides immediate feedback on algorithm performance. 
Starting with mathematical game theory research, the most common technique for game AI has been brute force search of the tree of legal moves. 
This technique, highly tuned and optimized, is what led to Chinook's ascendance to World Champion in 1994 and Deep Blue's eventual victory in 1997 over then World Champion Gary Kasparov. These victories cemented brute force tree search as standard techniques in abstract strategy games, to the point where computers constructing endgame tables have been strong enough to find sequences lasting for hundreds of moves. Kasparov himself suggested a version of Chess that allowed the player to consult a computer for move advice. Chinook went so far as to solve checkers, proving that perfect play will end in a draw. 
Several games, such as  Checkers, Chess, and Othello, have good enough evaluation functions and few enough legal moves so that that brute force tree search techniques can beat the best human players.

@@

Our experiments answer several questions, including: \\
$\bullet$  How beneficial are the learned policies ?  \\
$\bullet$   How stable is the evolutionary learning process? \\
$\bullet$   Do the learned policies work on different board sizes?

@@

\begin{figure}[tb]
\resizebox{3.5in}{!}{	\begin{HexBoard}[board size=13]
		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,h9,h5,h4,i4,g9,f9,g8,f7,f8,e8}
	\end{HexBoard}
	}
	\caption{13x13 Hex Game, \hwhite{} to play.}
	\label{fig:13x13inprogress}
\end{figure}

\begin{figure}[tb]
\resizebox{3.5in}{!}{	\begin{HexBoard}[board size=13]
		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,
				   h9,h5,h4,i4,g9,f9,g8,f7,f8,
				   e8,e9,c10,k8,c9,g6,g7,c11,
				   d10,h6,h7,i5,j5,j4,k4,k3,
				   m2,l3,m3,l4,c6,e11,l5,m4,
				   d11,b7,e6,d12,c12,c7,e7,f5,
				   e5,d9,d8,f4,e4,f3,a2,e3,d6,
				   c3,d3,d4,b5,e2,d2,d1,e1,b13,
				   c13,c5,b6}
	\end{HexBoard}
	}
	\caption{Finished 13x13 Hex Game, \hblack{}'s play 71 (at \loc{b6}) gives him an unbroken path from top to bottom.}
	\label{fig:13x13finished}
\end{figure}

@@

\begin{figure}
	\begin{center}
	\includegraphics[width=3.0in]{graphics/tree.pdf}
	\end{center}
	\caption{Partial MC Search Tree}
	\label{fig:tree}
\end{figure}

@@

\subsection{Child Selection Policy}
The \verb+select_child+ function defines the child selection policy. It must return the child of the given node to step into next. 
In general, child selection must trade off between \emph{exploitation} and \emph{exploration}. 
It is important to ``exploit" those child nodes with high win-rates, both
to discover possible weaknesses and to obtain highly accurate win-rate estimates. 
Simultaneously, a child with a low number of visits must also be explored since a low win-rate
could be due to a few unlucky simulations.
This exploration/exploitation tradeoff is related to a variance reduction problem.  
We want very low variance on the estimates for the good moves in order to maximize the chances of making the best move.
On the other hand, if the variance of a seemingly poor move is very high, then there is a significant chance that that move
could actually be very good.

@@

\begin{figure}
	\begin{center}
	\includegraphics[width=3.0in]{graphics/local-pattern.pdf}
	\caption{A 4x4 board with the 3 patterns shown around \hblacks\ last move. Example weights are shown in the center vertex.}
	\label{fig:localpattern}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
	\includegraphics[width=3.0in]{graphics/weight-pattern-map.pdf}
	\caption{The encoding from pattern A to a 32-bit integer. Using the weight from figure \ref{fig:localpattern}, the real value 6 would be stored at position 102 in the weight map.}
	\label{fig:encoding}
	\end{center}
\end{figure}

@@

%There are three simple continuation policies that we use as baseline comparators for our learned policies.
In principle any player could be used to evaluate the effectiveness of learning.   We used three simple continuation policies that were easily implementable in Hivemind as baseline comparators for our learned policies.
% TODO 9
% comment: would it be possible to include patterns/weights from competition-level programs such as MoGo (in Go)?
% answer: yes
% DPH: the more I think about it the more I think I don't understand the comment.  Is it about representing MoGo patterns with the 
% pattern encodings, or about using mogo patterns as a comparator?

@@

The Game Tracker module uses an interface that is easily extendible to other games. 
In addition to the \texttt{Playout} function, the Game Tracker provides the following functionality.
\begin{enumerate}
\item A single function, \texttt{Play(color, vertex)}, is used by the search tree to make a move in the simulated game.
\item The \texttt{Legal(color, vertex)} function is used to query the legality of a given move in the simulated game. 
The MCTS module uses this when the tree is gown as each time a leaf node is expanded,
one child is added for each legal move from the leaf node's state.
\item The \texttt{Winner()} function is used to determine the current winner of the game. 
	A special \texttt{Empty} player is returned if the game has not finished.
\item The \texttt{WasPlayed(color, vertex)} function provides information required for the AMAF heuristic.
\end{enumerate}
A few extra functions also provide for debugging by pretty-printing the tracker's internal state.

@@

\subsubsection*{Benefits of learned policy}
To determine the benefits of the learned continuation policy we played a round-robin set of matches between the learned, 
default, uniform local, and uniform with tenuki policies.
Each match consisted of 200 games on a 7x7 board where the first player alternated was forced to start with a
swap-safe\footnote{A swap-safe first move is a neutral or mediocre move by the first player in an attempt to eliminate the first-player advantage.  This is similar to the \emph{Komi} rule in Go.  We used C4 in our 7x7 games and C8 in the 11x11 games.}
move.
Each player was allowed 10,000 simulations per move and the results are shown in Figure~\ref{fig:results}.

@@

>>>>

\documentclass{acm_proc_article-sp}
%\documentclass{sig-alternate}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{url}
\usepackage{havannah}
\usepackage{enumitem}

\newcommand{\hblack}{\texttt{Black}}
\newcommand{\hwhite}{\texttt{White}}
\newcommand{\hblacks}{\texttt{Black's}}
\newcommand{\hwhites}{\texttt{White's}}
\newcommand{\loc}[1]{\texttt{#1}}

@@

\toappear{Permission to make digital or hard copies of all or part of this work for
   personal or classroom use is granted without fee provided that copies are
   not made or distributed for profit or commercial advantage and that copies
   bear this notice and the full citation on the first page. To copy otherwise,
   to republish, to post on servers or to redistribute to lists, requires prior
   specific permission and/or a fee. \\
   FDG '12, May 29-June 1, 2012 Raleigh, NC, USA. \\
   Copyright (c) 2012 ACM 978-1-4503-1333-9/12/05... \$10.00.}

@@

\setlist{topsep=0.5pt,partopsep=0.5pt,parsep=0.5pt,itemsep=0.5pt}

@@

\category{I.2.6}{Computing Methodologies}{Artificial Intelligence}

@@

\section{Introduction}

Abstract strategy games provide an attractive testbed for artificial intelligence and machine learning applications. The game framework provides immediate feedback on algorithm performance. 
Starting with mathematical game theory research, the most common technique for game AI has been brute force search of the tree of legal moves. 
This technique, highly tuned and optimized, is what led to Chinook's ascendance to World Champion in 1994 and Deep Blue's eventual victory in 1997 over then World Champion Gary Kasparov. These victories cemented brute force tree search as standard techniques in abstract strategy games, to the point where computers constructing endgame tables have been strong enough to find sequences lasting for hundreds of moves. Kasparov himself suggested a version of Chess that allowed the player to consult a computer for move advice. Chinook went so far as to solve checkers, proving that perfect play will end in a draw. 
Several games, such as  Checkers, Chess, and Othello, have good enough evaluation functions and few enough legal moves so that that brute force tree search techniques can beat the best human players.

@@

Our experiments answer several questions, including: 
\begin{itemize}
\item How beneficial are the learned policies ? 
\item How stable is the evolutionary learning process? 
\item Do the learned policies work on different board sizes?
\end{itemize}

@@

\begin{figure}[tb]
\resizebox{3.35in}{!}{	\begin{HexBoard}[board size=13]
		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,h9,h5,h4,i4,g9,f9,g8,f7,f8,e8}
	\end{HexBoard}
	}
	\caption{13x13 Hex Game, \hwhite{} to play.}
	\label{fig:13x13inprogress}
\end{figure}

\begin{figure}[tb]
\resizebox{3.35in}{!}{	\begin{HexBoard}[board size=13]
		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,
				   h9,h5,h4,i4,g9,f9,g8,f7,f8,
				   e8,e9,c10,k8,c9,g6,g7,c11,
				   d10,h6,h7,i5,j5,j4,k4,k3,
				   m2,l3,m3,l4,c6,e11,l5,m4,
				   d11,b7,e6,d12,c12,c7,e7,f5,
				   e5,d9,d8,f4,e4,f3,a2,e3,d6,
				   c3,d3,d4,b5,e2,d2,d1,e1,b13,
				   c13,c5,b6}
	\end{HexBoard}
	}
	\caption{Finished 13x13 Hex Game, \hblack{}'s play 71 (at \loc{b6}) gives him an unbroken path from top to bottom.}
	\label{fig:13x13finished}
\end{figure}

@@

\begin{figure}
	\begin{center}
	\includegraphics[width=2.5in]{graphics/tree.pdf}
	\end{center}
	\caption{Partial MC Search Tree}
	\label{fig:tree}
\end{figure}

@@

\subsection{Child Selection Policy}
The \verb+select_child+ function defines the child selection policy. It must return the child of the given node to step into next. 
In general, child selection must trade off between \emph{exploitation} and \emph{exploration}. 
It is important to ``exploit'' those child nodes with high win-rates, both
to discover possible weaknesses and to obtain highly accurate win-rate estimates. 
Simultaneously, a child with a low number of visits must also be explored since a low win-rate
could be due to a few unlucky simulations.
This exploration/exploitation tradeoff is related to a variance reduction problem.  
We want very low variance on the estimates for the good moves in order to maximize the chances of making the best move.
On the other hand, if the variance of a seemingly poor move is very high, then there is a significant chance that that move could actually be very good.

@@

\begin{figure}
	\begin{center}
	\includegraphics[width=2.85in]{graphics/local-pattern.pdf}
	\caption{A 4x4 board with the 3 patterns shown around \hblacks\ last move. Example weights are shown in the center vertex.}
	\label{fig:localpattern}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
	\includegraphics[width=3.25in]{graphics/weight-pattern-map.pdf}
	\caption{The encoding from pattern A to a 32-bit integer. Using the weight from figure \ref{fig:localpattern}, the real value 6 would be stored at position 102 in the weight map.}
	\label{fig:encoding}
	\end{center}
\end{figure}

@@

Note that many kinds of local hand-coded patterns (like those used by MoGo \cite{gelly2006modification} for Go) can be expressed using the Hivemind encoding system.

@@

The Game Tracker module uses an interface that is easily extendible to other games. 
In addition to the \texttt{Playout} function, the Game Tracker provides the following functionality.

\begin{enumerate}
\item A single function, \texttt{Play(color, vertex)}, is used by the search tree to make a move in the simulated game.
\item The \texttt{Legal(color, vertex)} function is used to query the legality of a given move in the simulated game. 
The MCTS module uses this when the tree is gown as each time a leaf node is expanded,
one child is added for each legal move from the leaf node's state.
\item The \texttt{Winner()} function is used to determine the current winner of the game. 
	A special \texttt{Empty} player is returned if the game has not finished.
\item The \texttt{WasPlayed(color, vertex)} function provides information required for the AMAF heuristic.
\end{enumerate}

A few extra functions also provide for debugging by pretty-printing the tracker's internal state.

@@

\subsubsection*{Benefits of learned policy}
In principle any player could be used to evaluate the effectiveness of learning.   
We used a round-robin tournament between the learned policy and the three following continuation policies that were easily implementable in Hivemind.

Each match consisted of 200 games on a 7x7 board where the first player alternated was forced to start with a
swap-safe\footnote{A swap-safe first move is a neutral or mediocre move by the first player in an attempt to eliminate the first-player advantage.  This is similar to the \emph{Komi} rule in Go.  We used C4 in our 7x7 games and C8 in the 11x11 games.}
move.
Each player was allowed 10,000 simulations per move and the results are shown in Figure~\ref{fig:results}.

@@

\subsection*{Acknowledgements}
We thank David Doshay and the Hierarchical Systems Research Foundation for the use of the their compute cluster. \\
